---
title: "xsynthdid"
author: Sebastian Kranz, Ulm University
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(xsynthdid)
```

### Overview

[Arkhangelsky et al. (2021)](https://www.aeaweb.org/articles?id=10.1257/aer.20190159) introduce the novel synthetic difference-in-differences (SDID) method that is implemented in the R package [synthdid](https://github.com/synth-inference/synthdid).

This package has the function `adjust.outcome.for.x` that can be used to adjust the outcome for exogenous time-varying covariates. For more background see my vignette [Synthetic Difference-in-Differences with Time-Varying Covariates (2021)](https://github.com/skranz/xsynthdid/blob/main/paper/synthdid_with_covariates.pdf).

### Installation

You can install the package from my [r-universe repository](https://skranz.r-universe.dev/ui#builds) by using the following code:

```{r eval=FALSE}
options(repos = c(skranz = 'https://skranz.r-universe.dev',
    CRAN = 'https://cloud.r-project.org'))
install.packages("xsynthdid")
```

### Usage example

We first simulate a panel data set with a true treatment effect of `tau=50` using the function `xsdid.mc` in our package.
```{r}
set.seed(1)
library(xsynthdid)
dat = xsdid.mc(N=30, T=30,tau=50,return.data = TRUE)
head(dat)
```

We first perform SDID estimation without any adjustment for the exogenous covariate `x`:
```{r}
pm = panel.matrices(dat, unit="i",time = "t",outcome = "y",treatment = "treat_exp")
sdid = synthdid_estimate(Y=pm$Y,N0=pm$N0,T0 = pm$T0) # inconsistent
sdid
```

We see how the estimator is a bit off the true causal effect.

We now adjust the outcomes `y` using the `adjust.outcome.for.x` function and use the adjusted outcome for the SDID estimation.
```{r}
dat$y.adj = adjust.outcome.for.x(dat,unit="i",time = "t",
  outcome = "y",treatment = "treat_exp", x="x")
pm = panel.matrices(dat, unit="i",time = "t",outcome = "y.adj",treatment = "treat_exp")
sdid.adj = synthdid_estimate(Y=pm$Y,N0=pm$N0,T0 = pm$T0) # inconsistent
sdid.adj
```

Our estimate is now pretty close to the true causal effect.
